---
title: "An introduction to Refate package"
author:
- name: Di Xiao
  affiliation:
  - Computational Systems Biology Group, Children’s Medical Research Institute, Faculty of Medicine and Health, The University of Sydney
- name: Pengyi Yang
  affiliation:
  - School of Mathematics and Statistics, The University of Sydney
  - Computational Systems Biology Group, Children’s Medical Research Institute, Faculty of Medicine and Health, The University of Sydney
  - Charles Perkins Centre, The University of Sydney
output:
  BiocStyle::html_document:
    toc_newpage: true
    fig_retina: NULL
package: BiocStyle
vignette: >
  %\VignetteIndexEntry{An introduction to Refate package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Installation

Install the latest development version from GitHub using the `devtools` package:

```{r, eval = FALSE}
library(devtools)
devtools::install()    # install the package
```

# Loading packages and data
```{r}
suppressMessages(
  {
    library(Refate)
    library(igraph)
    library(tidyverse)
    library(parallel)
    library(networkD3)
  }
)

# Load precomputed CPS list and TF-target interaction data
data(CPSs)    # Precomputed CPS values for human and mouse
data(TFtarget)    # TF-target interaction network
data(drugDB.weight)  # Drug-target interaction data
data(TFs_Kinases)
data(StringDB)
data(drugbases_metas)
data(target_data)
```

# Calculate Cellular Conversion Score (CCS)
```{r}
# Example log2 fold-change values (replace with your own)
data(fc.fibroblast2hiPSC)

# Calculate CCS for human genes
CCS <- calcCCS(fc.fibroblast2hiPSC, specie = "human")

# View the top CCS scores
head(sort(CCS, decreasing = TRUE))

# Plot the CCS for known TFs
tf.fibroblast2hiPSC = unique(c("POU5F1", "SOX2", "KLF4", "MYC", 
                      "OCT4", "SOX2", "NANOG", "LIN28", # https://doi.org/10.1126/science.1151526
                      "ESRRB", "UTF1", "LIN28B", "DPPA2", # https://doi.org/10.1016/j.cell.2012.08.023
                      "GLIS1", # https://doi.org/10.1038/nature10106
                      "NR5A2", # https://doi.org/10.1016/j.stem.2009.12.009
                      "MYCN",
                      "HMGA1"
                      ))

p <- rank.plot(CCS, gt = tf.fibroblast2hiPSC)
p
```

# Prioritize Drugs
```{r, fig.width=2, fig.height=4, warning=FALSE}

# Prioritize drugs
DPS <- calcDrugScore(
  CCS,
  community_cutoff = 10,
  community_cutoff2 = 600,
  filtering = TRUE,
  drugTarget_cutoff = 500,
  TFWeight = 2
)

# View the top-ranked drugs
head(DPS)

```


# Build the GRNs
We can build the transcriptional regulatory networks (GRNs) for a given drug. For example, we can build the GRNs for the known drug "Quercetin" by using the following code:

```{r}
GRNs <- build_GRNs(
  drug = "luteolin", 
  CCS, 
  kinome, 
  TFs, 
  StringDB, 
  TFtarget, 
  CCS_threshold = 0.02 # the threshold for selecting genes within the TRN
)

GRNs

```

# Get drug annotations
We can get drug information from different drug databases

```{r}
drug_name <- get_drug_name("luteolin")
drug_name 
drug_annotations <- get_drug_annotations("luteolin")
drug_annotations
drug_description <- get_drug_description("luteolin")
drug_description
```

# Filtering drugs based on toxicity

filter drugs based on the toxicity information from PubChem database
```{r, warning=FALSE}
## we first identify the top drugs
top_drugs <- names(DPS)[which(DPS > 0.9) ]
print(paste0("the number of top drugs: ", length(top_drugs)))
drug_names <- top_drugs[1:100] # only use the top 100 drugs for demo

data(all_drug_cids) 
out <- filter_toxicity(drug_names, 
                       all_drug_cids = all_drug_cids,
                       json_dir = "tmp/public/PubChem_json")

out$resolution  # check which names mapped to which CIDs
out$filtered_table %>% dplyr::filter(Keep) %>% dplyr::select(CID, query, MW)

```


# Clutering top drugs based on their target profiles
To characterize mechanistic diversity among top candidate drugs identified from the ESC-to-CNCC conversion analysis, we clustered drugs based on their gene target directionality profiles. This approach groups drugs with similar target engagement patterns and supports the rational selection of compounds spanning distinct mechanisms of action for downstream experimental validation.

```{r}

data(ESC2CNCC_top_drug)
data(target_data)
data(all_drug_cids)

cids <- all_drug_cids$cid[match(ESC2CNCC_top_drug, all_drug_cids$query)]

target_data_subset <- target_data %>%
  dplyr::filter(cid %in% cids) %>% 
  dplyr::select(cid, genename, direction) %>% 
  dplyr::filter(!is.na(genename))

mat <- target_data_subset %>%
  dplyr::group_by(genename, cid) %>%
  dplyr::summarise(direction = direction[1], .groups = "drop") %>%   # or mean(direction)
  pivot_wider(names_from = cid, values_from = direction, values_fill = 0) %>%
  tibble::column_to_rownames("genename") %>%
  as.matrix()

colnames(mat) <- ESC2CNCC_top_drug[match(colnames(mat), cids)]

# remove genes with no variation across drugs
mat_filt <- mat[apply(mat, 1, sd) > 0, , drop = FALSE]
mat_filt <- mat_filt[, apply(mat_filt, 2, sd) > 0, drop = FALSE]

# restrict to genes hit by at least k drugs:
k <- 10
mat_filt <- mat_filt[rowSums(mat_filt != 0) >= k, , drop = FALSE]

# Require that each drug hits at least m genes after filtering.
m <- 10  
mat_filt2 <- mat_filt[, colSums(mat_filt != 0) >= m, drop = FALSE]

# distance matrix
cor_drug <- cor(mat_filt, method = "pearson", use = "pairwise.complete.obs")
dist_drug <- as.dist(1 - cor_drug)

hc <- hclust(dist_drug, method = "average")  # average or ward.D2
plot(hc, main = "Drug clustering based on gene targeting", xlab = "", sub = "")

### heatmap visualisation
library(pheatmap)

pheatmap(
  t(mat_filt2),
  cluster_rows = hc,   # <- your precomputed tree
  cluster_cols = TRUE,      # or another hclust if you have one
  scale = "none",
  show_rownames = TRUE,
  show_colnames = FALSE,
  color = colorRampPalette(c("#5074AF", "lightgrey", "#C64032"))(100)
)
```

