---
title: "An introduction to Refate package"
author:
- name: Di Xiao
  affiliation:
  - Computational Systems Biology Group, Children’s Medical Research Institute, Faculty of Medicine and Health, The University of Sydney
- name: Pengyi Yang
  affiliation:
  - School of Mathematics and Statistics, The University of Sydney
  - Computational Systems Biology Group, Children’s Medical Research Institute, Faculty of Medicine and Health, The University of Sydney
  - Charles Perkins Centre, The University of Sydney
output:
  BiocStyle::html_document:
    toc_newpage: true
    fig_retina: NULL
package: BiocStyle
vignette: >
  %\VignetteIndexEntry{An introduction to Refate package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Installation

Install the latest development version from GitHub using the `devtools` package:

```{r, eval = FALSE}
library(devtools)
devtools::install()    # install the package
```

# Loading packages and data
```{r}
suppressMessages(
  {
    library(Refate)
    library(igraph)
    library(tidyverse)
    library(parallel)
  }
)

# Load precomputed CPS list and TF-target interaction data
data(CPSs)    # Precomputed CPS values for human and mouse
data(TFtarget)    # TF-target interaction network
data(drugDB.weight)  # Drug-target interaction data
data(TFs_Kinases)
data(StringDB)
data(drugbases_metas)
```

# Calculate Cellular Conversion Score (CCS)
```{r}
# Example log2 fold-change values (replace with your own)
data(fc.fibroblast2hiPSC)

# Calculate CCS for human genes
CCS <- calcCCS(fc.fibroblast2hiPSC, specie = "human")

# View the top CCS scores
head(sort(CCS, decreasing = TRUE))

# Plot the CCS for known TFs
tf.fibroblast2hiPSC = unique(c("POU5F1", "SOX2", "KLF4", "MYC", 
                      "OCT4", "SOX2", "NANOG", "LIN28", # https://doi.org/10.1126/science.1151526
                      "ESRRB", "UTF1", "LIN28B", "DPPA2", # https://doi.org/10.1016/j.cell.2012.08.023
                      "GLIS1", # https://doi.org/10.1038/nature10106
                      "NR5A2", # https://doi.org/10.1016/j.stem.2009.12.009
                      "MYCN",
                      "HMGA1"
                      ))

p <- rank.plot(CCS, gt = tf.fibroblast2hiPSC)
p
```

# Prioritize Drugs
```{r, fig.width=2, fig.height=4, warning=FALSE}

# Prioritize drugs
DPS <- calcDrugScore(
  CCS,
  community_cutoff = 10,
  community_cutoff2 = 600,
  filtering = TRUE,
  drugTarget_cutoff = 500,
  TFWeight = 2
)

# View the top-ranked drugs
head(DPS)

```


# Build the GRNs
We can build the transcriptional regulatory networks (GRNs) for a given drug. For example, we can build the GRNs for the known drug "Quercetin" by using the following code:

```{r}
GRNs <- build_GRNs(
  drug = "luteolin", 
  CCS, 
  kinome, 
  TFs, 
  StringDB, 
  TFtarget, 
  CCS_threshold = 0.02 # the threshold for selecting genes within the TRN
)

GRNs

```

# Get drug annotations
We can get drug information from different drug databases

```{r}
drug_name <- get_drug_name("luteolin")
drug_name 
drug_annotations <- get_drug_annotations("luteolin")
drug_annotations
drug_description <- get_drug_description("luteolin")
drug_description
```

# Filtering drugs based on toxicity

```{r, warning=FALSE}
## we first identify the top drugs
top_drugs <- names(DPS)[which(DPS > 0.9) ]
print(paste0("the number of top drugs: ", length(top_drugs)))
drug_names <- top_drugs[1:100] # only use the top 100 drugs for demo

data(all_drug_cids) 
out <- filter_toxicity(drug_names, 
                       all_drug_cids,
                       json_dir = "tmp/public/PubChem_json")

out$resolution  # check which names mapped to which CIDs
out$filtered_table %>% dplyr::filter(Keep) %>% dplyr::select(CID, query, MW)

```


