---
title: "An introduction to Refate package"
author:
- name: Di Xiao
  affiliation:
  - Computational Systems Biology Group, Children’s Medical Research Institute, Faculty of Medicine and Health, The University of Sydney
- name: Pengyi Yang
  affiliation:
  - School of Mathematics and Statistics, The University of Sydney
  - Computational Systems Biology Group, Children’s Medical Research Institute, Faculty of Medicine and Health, The University of Sydney
  - Charles Perkins Centre, The University of Sydney
output:
  BiocStyle::html_document:
    toc_newpage: true
    fig_retina: NULL
package: BiocStyle
vignette: >
  %\VignetteIndexEntry{An introduction to Refate package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Installation

Install the latest development version from GitHub using the `devtools` package:

```{r, eval = FALSE}
library(devtools)
devtools::install()    # install the package
```

# Loading packages and data
```{r}
suppressMessages(
  {
    library(Refate)
    library(igraph)
    library(tidyverse)
    library(parallel)
  }
)

# Load precomputed CPS list and TF-target interaction data
data(CPSs)    # Precomputed CPS values for human and mouse
data(TFtarget)    # TF-target interaction network
data(drugDB.weight)  # Drug-target interaction data
data(TFs_Kinases)
data(StringDB)
data(drugbases_metas)
```

# Calculate Cellular Conversion Score (CCS)
```{r}
# Example fold-change values (replace with your own)
data(fc.fibroblast2hiPSC)

# Calculate CCS for human genes
CCS <- calcCCS(fc.fibroblast2hiPSC, specie = "human")

# View the top CCS scores
head(sort(CCS, decreasing = TRUE))

# Plot the CCS for known TFs
tf.fibroblast2hiPSC = unique(c("POU5F1", "SOX2", "KLF4", "MYC", 
                      "OCT4", "SOX2", "NANOG", "LIN28", # https://doi.org/10.1126/science.1151526
                      "ESRRB", "UTF1", "LIN28B", "DPPA2", # https://doi.org/10.1016/j.cell.2012.08.023
                      "GLIS1", # https://doi.org/10.1038/nature10106
                      "NR5A2", # https://doi.org/10.1016/j.stem.2009.12.009
                      "MYCN",
                      "HMGA1"
                      ))

p <- rank.plot(CCS, gt = tf.fibroblast2hiPSC)
p
```

# Prioritize Drugs
```{r, fig.width=2, fig.height=4, warning=FALSE}

# Prioritize drugs
DPS <- calcDrugScore(
  CCS,
  community_cutoff = 10,
  community_cutoff2 = 600,
  filtering = TRUE,
  drugTarget_cutoff = 500,
  TFWeight = 2
)

# View the top-ranked drugs
head(DPS)

```


# build the TRNs
We can build the transcriptional regulatory networks (TRNs) for a given drug. For example, we can build the TRNs for the known drug "Quercetin" by using the following code:

```{r}
TRNs <- build_TRNs(
  drug = "quercetin", 
  CCS, 
  kinome, 
  TFs, 
  StringDB, 
  TFtarget, 
  CCS_threshold = 0.02 # the threshold for selecting genes within the TRN
)

TRNs
```

# Get drug annotations
We can get drug information from different drug databases

```{r}
drug_name <- get_drug_name("chir99021")
drug_name 
drug_annotations <- get_drug_annotations("chir99021")
drug_annotations
drug_description <- get_drug_description("chir99021")
drug_description
```

# Look for drugs with similar mechanisms of action with the known chemical compound
for example, we can find drugs with similar mechanisms of action with the known chemical compound "chir99021" by using the following code:

```{r, warning=FALSE}
## we first identify the top drugs
top_drugs <- names(DPS)[which(DPS > 0.9) ]
print(paste0("the number of top drugs: ", length(top_drugs)))
  
similar_drugs <- identify_similar_drugs(known_drug = "chir99021",
                                        top_drugs,
                                        CCS,
                                        drugDB.weight,
                                        method = "spearman",
                                        top = 30)
head(similar_drugs)
```


