---
title: "iPSC to CNCC image quantification"
author: "Di Xiao"
date: "`r paste0('Compiled ', format(Sys.time(), '%Y %b %d'))`"
output:
  html_document:
    code_folding: hide
    fig_height: 6
    fig_width: 6
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float: true
    theme: paper
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This notebook quantifies marker-positive cells from CellProfiler outputs for
the iPSC drug-screen validation. 

# Setup

```{r libs}
suppressPackageStartupMessages({
  library("readr")
  library("readxl")
  library("ggpubr")
  library("ggplot2")
  library("dplyr")
  library("tidyr")
})

scPalette <- function(n) {
  color_space <- c(
    "#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#F29403", "#F781BF",
    "#BC9DCC", "#A65628", "#54B0E4", "#222F75", "#1B9E77", "#B2DF8A",
    "#E3BE00", "#FB9A99", "#E7298A", "#910241", "#00CDD1", "#A6CEE3",
    "#CE1261", "#5E4FA2", "#8CA77B", "#00441B", "#DEDC00", "#B3DE69",
    "#8DD3C7", "#999999"
  )

  if (n <= length(color_space)) {
    color_space[seq_len(n)]
  } else {
    grDevices::colorRampPalette(color_space)(n)
  }
}
```

```{r config}
target_condition <- params$target_condition
control_name <- "CHIR"

# Positive-cell threshold: q = 0.6 retains the top 40% of per-cell intensities.
cutoff_quantile <- 0.6

# CellProfiler channel used for the lineage marker. Change this to "FITC" if
# the final staining/imaging run stored the marker in FITC rather than Cy5.
marker_channel <- "Cy5"
marker_label <- "SOX10"
```

# Annotation tables

```{r annotation_helpers}
clean_drug <- function(x) {
  x <- trimws(x)
  x <- gsub("\\s+", " ", x)
  x <- sub("([A-Za-z])([0-9]+(?:\\.[0-9]+)?[unm][Mm])$", "\\1 \\2", x)
  x <- sub("(?i)^nothing$", "Secondary", x, perl = TRUE)
  x <- sub("^Secondary.*", "Secondary", x)
  x
}

pad_well <- function(w) {
  m <- regmatches(w, regexec("^([A-P])([0-9]+)$", w))
  vapply(m, function(mm) {
    if (length(mm) == 3) sprintf("%s%02d", mm[2], as.integer(mm[3])) else NA_character_
  }, character(1))
}

clean_plate <- function(p) gsub("\\s+", "", as.character(p))

normalize_key <- function(df) {
  df %>%
    rename(Treatment_Dosage = Drug, Original_Repeat = Replicate) %>%
    mutate(
      Plate = clean_plate(Plate),
      Well = pad_well(Well),
      Condition = tolower(Condition),
      Treatment_Dosage = clean_drug(Treatment_Dosage)
    ) %>%
    filter(!is.na(Well))
}
```

```{r load_annotations}
key_new <- "20260408/20260421_new_key_iPSCEXP.xlsx"
key_rep5 <- "20260408/20260421_Di_imageanalysis_key.xlsx"

anno_rep123 <- read_xlsx(key_new, sheet = "20260326_iPSC_EXP") %>%
  select(Plate, Well, `Cell Line`, Replicate, Condition, Drug) %>%
  normalize_key()

anno_rep4 <- read_xlsx(key_rep5, sheet = "iPSC 2.5uM CHIR 5th replicate") %>%
  select(Plate, Well, `Cell Line`, Replicate, Condition, Drug) %>%
  normalize_key()
```

# Load CellProfiler outputs

```{r load_helpers}
required_columns <- function(df, cols, file_label) {
  missing <- setdiff(cols, names(df))
  if (length(missing) > 0) {
    stop("Missing required column(s) in ", file_label, ": ",
         paste(missing, collapse = ", "), call. = FALSE)
  }
}

get_image_results <- function(image_csv, anno, marker_channel) {
  image_df <- read.csv(image_csv, check.names = FALSE)

  marker_file_col <- paste0("FileName_", marker_channel)
  marker_total_col <- paste0(
    "Intensity_TotalIntensity_", marker_channel, "_DAPI_Cells_Filtered"
  )
  marker_mean_col <- paste0(
    "Intensity_MeanIntensity_", marker_channel, "_DAPI_Cells_Filtered"
  )
  marker_cell_mean_col <- paste0(
    "Mean_DAPI_Cells_Filtered_Intensity_MeanIntensity_", marker_channel
  )
  marker_cell_median_col <- paste0(
    "Mean_DAPI_Cells_Filtered_Intensity_MedianIntensity_", marker_channel
  )

  required_columns(
    image_df,
    c("ImageNumber", "FileName_DAPI", marker_file_col,
      "AreaOccupied_TotalArea_DAPI_Cells_Filtered",
      "AreaOccupied_AreaOccupied_DAPI_Cells_Filtered",
      "Count_DAPI_Cells_Filtered",
      "Mean_DAPI_Cells_Filtered_AreaShape_Area",
      marker_total_col, marker_mean_col,
      marker_cell_mean_col, marker_cell_median_col),
    image_csv
  )

  image_df %>%
    transmute(
      ImageNumber,
      Well = sapply(strsplit(FileName_DAPI, "_"), `[[`, 2),
      DAPI_FileName = FileName_DAPI,
      Marker_FileName = .data[[marker_file_col]],
      Total_Area = AreaOccupied_TotalArea_DAPI_Cells_Filtered,
      DAPI_Area = AreaOccupied_AreaOccupied_DAPI_Cells_Filtered,
      DAPI_Cells_Count = Count_DAPI_Cells_Filtered,
      DAPI_Cells_MeanArea = Mean_DAPI_Cells_Filtered_AreaShape_Area,
      Marker_TotalIntensity_in_DAPI = .data[[marker_total_col]],
      Marker_MeanIntensity_in_DAPI = .data[[marker_mean_col]],
      DAPI_Cells_Marker_MeanIntensity = .data[[marker_cell_mean_col]],
      DAPI_Cells_Marker_MedianIntensity = .data[[marker_cell_median_col]]
    ) %>%
    full_join(anno, by = "Well")
}

read_plate <- function(image_csv, cell_csv, anno, original_repeat,
                       reported_repeat, plate_id,
                       condition = target_condition,
                       marker_channel = marker_channel) {
  anno_plate <- anno %>%
    filter(Plate == plate_id, Original_Repeat == original_repeat) %>%
    mutate(Repeat = reported_repeat)

  per_image <- get_image_results(image_csv, anno_plate, marker_channel) %>%
    filter(Condition == condition, Repeat == reported_repeat)

  meta <- per_image %>%
    select(ImageNumber, Plate, Well, Repeat, Original_Repeat,
           Condition, Treatment_Dosage)

  per_cell <- read.csv(cell_csv, check.names = FALSE) %>%
    left_join(meta, by = "ImageNumber") %>%
    filter(Condition == condition, Repeat == reported_repeat)

  list(perImage = per_image, perCell = per_cell)
}
```

```{r load_all, warning=FALSE, message=FALSE}
plate_manifest <- list(
  list(reported_repeat = 1L, original_repeat = 2L, plate_id = "Plate_2935",
       image_csv = "20260325/p1_Image.csv",
       cell_csv = "20260325/p1_DAPI_Cells_Filtered.csv",
       anno = anno_rep234),
  list(reported_repeat = 2L, original_repeat = 3L, plate_id = "Plate_2936",
       image_csv = "20260325/p2_Image.csv",
       cell_csv = "20260325/p2_DAPI_Cells_Filtered.csv",
       anno = anno_rep234),
  list(reported_repeat = 3L, original_repeat = 4L, plate_id = "Plate_2937",
       image_csv = "20260325/p3_Image.csv",
       cell_csv = "20260325/p3_DAPI_Cells_Filtered.csv",
       anno = anno_rep234),
  list(reported_repeat = 4L, original_repeat = 5L, plate_id = "Plate_3045",
       image_csv = "20260408/p5_Image.csv",
       cell_csv = "20260408/p5_DAPI_Cells_Filtered.csv",
       anno = anno_rep5)
)

reps <- lapply(plate_manifest, function(x) {
  read_plate(
    image_csv = x$image_csv,
    cell_csv = x$cell_csv,
    anno = x$anno,
    original_repeat = x$original_repeat,
    reported_repeat = x$reported_repeat,
    plate_id = x$plate_id,
    marker_channel = marker_channel
  )
})
names(reps) <- paste0("rep", vapply(plate_manifest, `[[`, integer(1), "reported_repeat"))

for (nm in names(reps)) {
  cat(sprintf("%s  images=%4d  cells=%6d\n",
              nm, nrow(reps[[nm]]$perImage), nrow(reps[[nm]]$perCell)))
}
```

# Positive-cell thresholds

```{r thresholds, fig.height=3, fig.width=5}
marker_mean_col <- paste0("Intensity_MeanIntensity_", marker_channel)

thresholding <- function(per_cell, channel_col, q = cutoff_quantile) {
  required_columns(per_cell, channel_col, "per-cell CellProfiler table")
  thr <- quantile(per_cell[[channel_col]], q, na.rm = TRUE)
  hist(per_cell[[channel_col]], breaks = 20,
       main = sprintf("%s mean intensity (q=%.2f, threshold=%.3f)",
                      marker_channel, q, thr),
       xlab = paste(marker_label, "mean intensity per cell"))
  abline(v = c(thr, 1), col = "red")
  list(upper = 1, lower = thr)
}

thresholds <- lapply(reps, function(r)
  thresholding(r$perCell, marker_mean_col, cutoff_quantile))
```

# Per-replicate summaries

```{r calc_summary}
calc_intensity_summary <- function(per_cell, per_image, intensity_col,
                                   lower, upper,
                                   group_vars = c("Treatment_Dosage", "Repeat")) {
  required_columns(per_cell, intensity_col, "per-cell CellProfiler table")

  intensity_tbl <- per_cell %>%
    filter(.data[[intensity_col]] >= lower,
           .data[[intensity_col]] <= upper) %>%
    group_by(ImageNumber) %>%
    summarise(
      intensity_fil = mean(.data[[intensity_col]], na.rm = TRUE),
      total_intensity_fil = sum(.data[[intensity_col]], na.rm = TRUE),
      .groups = "drop"
    ) %>%
    complete(ImageNumber = unique(per_cell$ImageNumber),
             fill = list(intensity_fil = NA_real_,
                         total_intensity_fil = NA_real_))

  positive_tbl <- per_cell %>%
    group_by(ImageNumber) %>%
    summarise(
      Cell_number = n(),
      Positive_Cell_Proportion = sum(.data[[intensity_col]] > lower, na.rm = TRUE) / n(),
      Positive_Cell_Count = sum(.data[[intensity_col]] > lower, na.rm = TRUE),
      .groups = "drop"
    )

  per_image_results <- per_image %>%
    left_join(intensity_tbl, by = "ImageNumber") %>%
    left_join(positive_tbl, by = "ImageNumber")

  summaries <- per_image_results %>%
    group_by(across(all_of(group_vars))) %>%
    summarise(
      Positive_Cell_Proportion = mean(Positive_Cell_Proportion, na.rm = TRUE),
      Positive_Cell_Count = mean(Positive_Cell_Count, na.rm = TRUE),
      Marker_PositiveCell_MeanIntensity = mean(intensity_fil, na.rm = TRUE),
      Marker_PositiveCell_TotalIntensity = mean(total_intensity_fil, na.rm = TRUE),
      Cell_number = mean(Cell_number, na.rm = TRUE),
      .groups = "drop"
    )

  list(per_image_results = per_image_results,
       summaries_results = summaries)
}

outs <- mapply(function(r, t) {
  calc_intensity_summary(
    per_cell = r$perCell,
    per_image = r$perImage,
    intensity_col = marker_mean_col,
    lower = t$lower,
    upper = t$upper
  )
}, reps, thresholds, SIMPLIFY = FALSE)

combined_raw <- bind_rows(lapply(outs, `[[`, "summaries_results"))

drugs <- unique(combined_raw$Treatment_Dosage)
ordered_drugs <- c(control_name,
                   sort(setdiff(drugs, c(control_name, "Secondary"))),
                   intersect("Secondary", drugs))
combined_raw$Treatment_Dosage <- factor(combined_raw$Treatment_Dosage,
                                         levels = ordered_drugs)

cat("Treatments:", paste(levels(combined_raw$Treatment_Dosage), collapse = ", "), "\n")
cat("n per treatment:\n")
print(table(combined_raw$Treatment_Dosage))
```

# CHIR-anchored normalisation

Each reported replicate is one physical plate with its own CHIR control. The
analysis rescales each replicate by:

```text
value_Norm = value * (global_CHIR_mean / per_repeat_CHIR_mean)
```

This preserves within-plate fold-changes while aligning the absolute CHIR level
across replicates.

```{r batch_correct}
batch_correct_by_control <- function(data,
                                     batch_var = "Repeat",
                                     control_label = "CHIR",
                                     treatment_var = "Treatment_Dosage",
                                     measure_vars) {
  control_global <- data %>%
    filter(.data[[treatment_var]] == control_label) %>%
    summarise(across(all_of(measure_vars),
                     ~ mean(.x, na.rm = TRUE),
                     .names = "{.col}_CTavg"))

  control_batch <- data %>%
    filter(.data[[treatment_var]] == control_label) %>%
    group_by(.data[[batch_var]]) %>%
    summarise(across(all_of(measure_vars),
                     ~ mean(.x, na.rm = TRUE),
                     .names = "{.col}_BTavg"),
              .groups = "drop")

  out <- data %>%
    bind_cols(control_global[rep(1, nrow(data)), , drop = FALSE]) %>%
    left_join(control_batch, by = setNames(batch_var, batch_var))

  for (v in measure_vars) {
    ct_col <- paste0(v, "_CTavg")
    bt_col <- paste0(v, "_BTavg")
    fac_col <- paste0(v, "_NormFac")
    norm_col <- paste0(v, "_Norm")
    out[[fac_col]] <- out[[ct_col]] / out[[bt_col]]
    out[[norm_col]] <- out[[v]] * out[[fac_col]]
  }

  out
}

measure_vars <- c("Positive_Cell_Proportion",
                  "Positive_Cell_Count",
                  "Marker_PositiveCell_MeanIntensity",
                  "Marker_PositiveCell_TotalIntensity",
                  "Cell_number")

combined <- batch_correct_by_control(
  combined_raw,
  batch_var = "Repeat",
  control_label = control_name,
  measure_vars = measure_vars
)

cat("Per-replicate CHIR scaling factors for positive-cell proportion:\n")
print(combined %>%
        filter(Treatment_Dosage == control_name) %>%
        select(Repeat, Positive_Cell_Proportion,
               Positive_Cell_Proportion_NormFac,
               Positive_Cell_Proportion_Norm) %>%
        distinct())
```

# Statistical testing

One-sided Wilcoxon rank-sum tests compare each treatment with CHIR on the
normalised metrics. `alternative = "greater"` tests whether the treatment
produces a higher marker-positive signal than CHIR.

```{r stat_test}
run_wilcox <- function(df, metric, control = "CHIR", alternative = "greater") {
  treatments <- setdiff(levels(df$Treatment_Dosage), control)
  ctrl_vals <- df %>%
    filter(Treatment_Dosage == control) %>%
    pull(all_of(metric))

  bind_rows(lapply(treatments, function(trt) {
    trt_vals <- df %>%
      filter(Treatment_Dosage == trt) %>%
      pull(all_of(metric))

    base <- data.frame(
      Treatment_Dosage = trt,
      metric = metric,
      mean_ctrl = mean(ctrl_vals, na.rm = TRUE),
      mean_trt = mean(trt_vals, na.rm = TRUE),
      n_ctrl = sum(!is.na(ctrl_vals)),
      n_trt = sum(!is.na(trt_vals)),
      p.value = NA_real_
    )

    if (base$n_ctrl < 2 || base$n_trt < 2) return(base)

    wt <- suppressWarnings(
      wilcox.test(trt_vals, ctrl_vals,
                  alternative = alternative,
                  exact = FALSE)
    )
    base$p.value <- wt$p.value
    base
  }))
}

metrics_of_interest <- c("Positive_Cell_Proportion_Norm",
                         "Positive_Cell_Count_Norm",
                         "Marker_PositiveCell_MeanIntensity_Norm",
                         "Cell_number_Norm")

stat_results <- bind_rows(lapply(metrics_of_interest, function(m)
  run_wilcox(combined, m, control = control_name, alternative = "greater"))) %>%
  group_by(metric) %>%
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup() %>%
  mutate(
    sig_label = case_when(
      p.adj < 0.001 ~ "***",
      p.adj < 0.01 ~ "**",
      p.adj < 0.05 ~ "*",
      p.adj < 0.1 ~ ".",
      TRUE ~ "ns"
    ),
    fold_change = mean_trt / mean_ctrl
  ) %>%
  arrange(metric, p.adj)

print(stat_results %>%
        select(metric, Treatment_Dosage, n_ctrl, n_trt, mean_ctrl, mean_trt,
               fold_change, p.value, p.adj, sig_label),
      n = Inf)
```

# Plots

```{r normalised_plots, fig.height=6, fig.width=16}
n_trt <- length(levels(combined$Treatment_Dosage))
pal <- scPalette(n_trt)
comparisons <- lapply(setdiff(levels(combined$Treatment_Dosage), control_name),
                      function(t) c(control_name, t))

box_vs_ctrl <- function(df, y, ylab, title) {
  ggboxplot(df, x = "Treatment_Dosage", y = y,
            color = "Treatment_Dosage", add = "jitter", palette = pal) +
    stat_compare_means(
      comparisons = comparisons,
      method = "wilcox.test",
      # Comparisons are ordered as c(CHIR, treatment), so "less" tests
      # CHIR < treatment, equivalent to treatment > CHIR.
      method.args = list(alternative = "less", exact = FALSE),
      label = "p.signif",
      hide.ns = TRUE
    ) +
    xlab("Treatment") +
    ylab(ylab) +
    ggtitle(title) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(hjust = 0.5)
    )
}

p_prop <- box_vs_ctrl(
  combined,
  "Positive_Cell_Proportion_Norm",
  paste(marker_label, "positive cell proportion (CHIR-normalised)"),
  "iPSC replicates 1-4 - positive cell proportion"
)
p_count <- box_vs_ctrl(
  combined,
  "Cell_number_Norm",
  "Total cell number (CHIR-normalised)",
  "iPSC replicates 1-4 - cell number"
)
p_int <- box_vs_ctrl(
  combined,
  "Marker_PositiveCell_MeanIntensity_Norm",
  paste(marker_label, "mean intensity in positive cells (CHIR-normalised)"),
  paste("iPSC replicates 1-4 -", marker_label, "intensity")
)

ggarrange(p_prop, p_count, p_int, ncol = 3)
```

```{r raw_plots, fig.height=6, fig.width=16}
pal_rep <- scPalette(length(unique(combined_raw$Repeat)))

box_raw <- function(y, ylab, title) {
  df <- combined_raw %>% mutate(Repeat = factor(Repeat))

  ggplot(df, aes(x = Treatment_Dosage, y = .data[[y]])) +
    geom_boxplot(aes(color = Treatment_Dosage),
                 outlier.shape = NA,
                 width = 0.6) +
    geom_jitter(aes(color = Repeat),
                width = 0.15,
                size = 2,
                alpha = 0.9) +
    scale_color_manual(
      values = c(
        setNames(scPalette(length(levels(df$Treatment_Dosage))),
                 levels(df$Treatment_Dosage)),
        setNames(pal_rep, levels(df$Repeat))
      ),
      breaks = levels(df$Repeat),
      name = "Repeat"
    ) +
    stat_compare_means(
      comparisons = comparisons,
      method = "wilcox.test",
      method.args = list(alternative = "less", exact = FALSE),
      label = "p.signif",
      hide.ns = TRUE
    ) +
    xlab("Treatment") +
    ylab(ylab) +
    ggtitle(title) +
    theme_classic() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(hjust = 0.5)
    )
}

p_raw_prop <- box_raw(
  "Positive_Cell_Proportion",
  paste(marker_label, "positive cell proportion (raw)"),
  "iPSC replicates 1-4 - positive cell proportion (raw)"
)
p_raw_count <- box_raw(
  "Cell_number",
  "Cell number per site (raw)",
  "iPSC replicates 1-4 - cell number (raw)"
)
p_raw_int <- box_raw(
  "Marker_PositiveCell_MeanIntensity",
  paste(marker_label, "mean intensity in positive cells (raw)"),
  paste("iPSC replicates 1-4 -", marker_label, "intensity (raw)")
)

ggarrange(p_raw_prop, p_raw_count, p_raw_int, ncol = 3)
```

# Export tables

```{r export}
dir.create("output", showWarnings = FALSE)

marker_stub <- gsub("[^A-Za-z0-9]+", "_", marker_label)
out_base <- sprintf("output/iPSC_rep1to4_%s_%s", marker_stub, target_condition)

write_csv(combined, paste0(out_base, "_combined_summaries_CHIRnorm.csv"))
write_csv(stat_results, paste0(out_base, "_wilcox_vs_CHIR_normalised.csv"))

threshold_table <- tibble::tibble(
  Repeat = vapply(plate_manifest, `[[`, integer(1), "reported_repeat"),
  Original_Repeat = vapply(plate_manifest, `[[`, integer(1), "original_repeat"),
  Plate = vapply(plate_manifest, `[[`, character(1), "plate_id"),
  cutoff_q = cutoff_quantile,
  Marker_Channel = marker_channel,
  Marker_Label = marker_label,
  Marker_Lower = vapply(thresholds, `[[`, numeric(1), "lower"),
  Marker_Upper = vapply(thresholds, `[[`, numeric(1), "upper")
)
write_csv(threshold_table, paste0(out_base, "_per_replicate_thresholds.csv"))

percell_export <- bind_rows(lapply(seq_along(reps), function(i) {
  r <- reps[[i]]
  meta <- r$perImage %>%
    select(ImageNumber, Plate, Well, Repeat, Original_Repeat,
           Condition, Treatment_Dosage)

  r$perCell %>%
    select(ImageNumber, ObjectNumber,
           all_of(marker_mean_col),
           any_of(c(paste0("Intensity_MedianIntensity_", marker_channel),
                    paste0("Intensity_IntegratedIntensity_", marker_channel)))) %>%
    left_join(meta, by = "ImageNumber") %>%
    select(Plate, Repeat, Original_Repeat, Condition, Well, Treatment_Dosage,
           ImageNumber, ObjectNumber, everything())
}))

percell_path <- paste0(out_base, "_per_cell_", marker_channel, "_", marker_stub, ".csv.gz")
write_csv(percell_export, gzfile(percell_path))

cat("Wrote:\n",
    paste0(out_base, "_combined_summaries_CHIRnorm.csv"), "\n",
    paste0(out_base, "_wilcox_vs_CHIR_normalised.csv"), "\n",
    paste0(out_base, "_per_replicate_thresholds.csv"), "\n",
    percell_path, "\n")

cat(sprintf("Per-cell export dimensions: %d rows x %d columns\n",
            nrow(percell_export), ncol(percell_export)))
```

# Session information

```{r session_info}
sessionInfo()
```
